---
title: "CodeBook"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# CodeBook

## Data Set

This data set was generated out of the [Human Activity Recognition Using Smartphones Data Set](http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones) 
containing the recordings of 30 subjects performing activities of daily living (ADL) 
while carrying a waist-mounted smartphone with embedded inertial sensors.

It can be found in the file 'average_data.txt' and is generated by 
running the R script 'run_analysis.R'. The file contains a table with the values
separated by a blank and can be loaded with the following command:

```{r eval=FALSE}
dataSet <- read.table('average_data.txt', header = TRUE, sep = " ", stringsAsFactors = TRUE)
```

The resulting data frame consists of 4 columns:

1. subject: the numeric id of the person that generated the data
2. activity: the activity being performed (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING)
3. feature: the name of the variable in the original data set (66 features)
4. average: the calculated average of each feature grouped by subject and activity

```{r echo=FALSE}
library(dplyr)
dataSet <- read.table('average_data.txt', header = TRUE, sep = " ", stringsAsFactors = TRUE)
dataSet <- mutate(dataSet, subject = as.factor(subject))
```

```{r echo=TRUE}
summary(dataSet)
```
